---
title: "Class11"
author: "Wenxi Tang"
format: gfm
editor: visual
---

```{r}
#read the genotype csv file
gt <- read.csv("373531-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378.csv")
table(gt$Genotype..forward.strand.)/nrow(gt)
```

```{r}
gt$Genotype..forward.strand.
```

# Population scale analysis

Let's look if there is any association of the 4 asthma-associated SNPs on ORMDL3 expression

```{r}
url <- "https://bioboot.github.io/bggn213_W19/class-material/rs8067378_ENSG00000172057.6.txt"
```

```{r}
results <- read.table(url)
head(results)
```

```{r}
table(results$geno)
```

> Q13: Read this file into R and determine the sample size for each genotype and their corresponding median expression levels for each of these genotypes.

```{r}
inds <- results$geno == "G/G"
summary(results$exp[inds])
```

```{r}
inds <- results$geno == "A/A"
summary(results$exp[inds])
```

```{r}
inds <- results$geno == "A/G"
summary(results$exp[inds])
```

> Q14. Generate a boxplot with a box per genotype, what could you infer from the relative expression value between A/A and G/G displayed in this plot? Does the SNP effect the expression of ORMDL3?

expression: A/A seems to have highest expression compared to A/G then compared to G/G.

```{r}
library(ggplot2)

ggplot(results) +
  aes(geno, exp) +
  geom_boxplot(notch = T) +
  geom_jitter(width = 0.2, alpha = 0.2) +
  theme_bw()
```
