---
title: "Class17: Vaccination Rate Mini Project"
author: "Wenxi Tang"
format: gfm
editor: visual
---

##Background

In today's class, we will explore a dataset on state wide vaccination rate from CA.gov.

The goal of this hands-on mini-project is to examine and compare the Covid-19 vaccination rates around San Diego.

We will start by downloading the most recently dated "Statewide COVID-19 Vaccines Administered by ZIP Code" CSV file from: https://data.ca.gov/dataset/covid-19-vaccine-progress-dashboard-data-by-zip-code

## Data input

```{r}
# Import vaccination data
vax <- read.csv("covid19vaccinesbyzipcode_test.csv")
head(vax)
```

```{r}
library(skimr)
```

A useful function for exploring new datasets is **skimr** package.

```{r}
skimr::skim(vax)
```

```{r}
#sort(vax$as_of_date, decreasing = F)
#sort(vax$as_of_date)
```

> Q1. What column details the total number of people fully vaccinated?

The column "persons_fully_vaccinated"

> Q2. What column details the Zip code tabulation area?

The column "zip_code_tabulation_area"

> Q3. What is the earliest date in this dataset?

```{r}
vax$as_of_date[1]
```

2021-01-05

> Q4. What is the latest date in this dataset?

```{r}
vax$as_of_date[nrow(vax)]
```

2023-03-07

> Q5. How many numeric columns are in this dataset?

13

> Q6. Note that there are "missing values" in the dataset. How many NA values there in the persons_fully_vaccinated column?

```{r}
sum(is.na(vax$persons_fully_vaccinated))
```

16621

> Q7. What percent of persons_fully_vaccinated values are missing (to 2 significant figures)?

```{r}
round(16621/nrow(vax) * 100, 2)
```

> Q8. \[Optional\]: Why might this data be missing?

It might be some confidential federal areas, or at very early stages, data was not collected.

##Working with dates

We will use the *lubridate* package to help ease the pain of working with times and dates.

```{r}
library(lubridate)
```

```{r}
today()
```

> Q9. How many days have passed since the last update of the dataset?

```{r}
today() - ymd(vax$as_of_date[nrow(vax)])
```

I will convert the entire "as_of_date" column to be in lubridate format.

```{r}
vax$as_of_date <- ymd(vax$as_of_date)
```

```{r}
today() - vax$as_of_date[(nrow(vax))]
```

```{r}
today() - ymd("1997-4-20")
```

> Q10. How many unique dates are in the dataset (i.e. how many different dates are detailed)?

```{r}
length(unique(vax$as_of_date))
```

```{r}
library(dplyr)
n_distinct(vax$as_of_date)
```

## Working with zipcode

There are quite a few R packages that can help ease the pain of working with ZIP codes, we will try just one of the smaller ones **zipcodeR**.

```{r}
library(zipcodeR)
```

```{r}
geocode_zip('92037')
```

```{r}
zip_distance('92037','92109')
```

```{r}
#Calculate the distance between the centroids of any two ZIP codes in miles, e.g.
reverse_zipcode(c('92037', "92109", "92122", "94061") )
```

> Q. Find the best and worst ratio of "median household income" to "median_home_value" in San Diego county

> Q11. How many distinct zip codes are listed for San Diego County?

```{r}
sd_zip <- unique(vax$zip_code_tabulation_area[vax$county == "San Diego"])
length(sd_zip)
```

```{r}
sd.eco <- reverse_zipcode(sd_zip)
```

Now extract the best and worst "median household income" and "median_home_value"

```{r}
sort(sd.eco$median_household_income/sd.eco$median_home_value, decreasing = T)
```

```{r}
arrange(sd.eco, sd.eco$median_household_income, sd.eco$median_home_value)
```

```{r}
# Pull data for all ZIP codes in the dataset
#zipdata <- reverse_zipcode( vax$zip_code_tabulation_area )
```

## Focus on the San Diego area

> Q12. What San Diego County Zip code area has the largest 12 + Population in this dataset?

```{r}
sd <- filter(vax, county == "San Diego")
```

```{r}
ind <- which.max(sd$age12_plus_population)
sd[ind,]
```

> Q13. What is the overall average "Percent of Population Fully Vaccinated" value for all San Diego "County" as of "2023-03-07"?

```{r}
vax$as_of_date[nrow(vax)]
```

```{r}
thisdate <- filter(sd, as_of_date == "2023-03-07")
```

```{r}
mean(thisdate$percent_of_population_fully_vaccinated, na.rm = T)
```

> Q14. Using either ggplot or base R graphics make a summary figure that shows the distribution of Percent of Population Fully Vaccinated values as of "2023-02-28"?

```{r}
hist(thisdate$percent_of_population_fully_vaccinated)
```

```{r}
library(ggplot2)
```

```{r}
ggplot(thisdate) +
  aes(percent_of_population_fully_vaccinated) +
  geom_histogram(binwidth = 0.1) +
  theme_bw() +
  xlab("percent of population fully vaccinated") +
  ylab("Counts")
```

## Focus on UCSD/La Jolla

UCSD resides in the 92037 zip code area.

```{r}
ucsd <- filter(sd, zip_code_tabulation_area=="92037")
ucsd[1,]$age5_plus_population
```

> Q15. Using ggplot make a graph of the vaccination rate time course for the 92037 ZIP code area:

```{r}
ljplot <- ggplot(ucsd) +
  aes(as_of_date,
      percent_of_population_fully_vaccinated) +
  geom_point() +
  geom_line(group=1) +
  ylim(c(0,1)) +
  labs(x = "Date", y="Percent Vaccinated")

ljplot
```

## Comparing to similar sized areas

Let's return to the full dataset and look across every zip code area with a population at least as large as that of 92037 on as_of_date "2023-03-07".

```{r}
# Subset to all CA areas with a population as large as 92037
vax.36 <- filter(vax, age5_plus_population > 36144 &
                as_of_date == "2023-03-07")

head(vax.36)
```

> Q16. Calculate the mean "Percent of Population Fully Vaccinated" for ZIP code areas with a population as large as 92037 (La Jolla) as_of_date "2023-02-28". Add this as a straight horizontal line to your plot from above with the geom_hline() function?

```{r}
n_distinct(vax.36$zip_code_tabulation_area)
```

```{r}
mean(vax.36$percent_of_population_fully_vaccinated, na.rm = T)
```

```{r}
ljplot +
  geom_hline(yintercept = 0.7214936, col = "red", linetype = 5)
```

> Q17. What is the 6 number summary (Min, 1st Qu., Median, Mean, 3rd Qu., and Max) of the "Percent of Population Fully Vaccinated" values for ZIP code areas with a population as large as 92037 (La Jolla) as_of_date "2023-03-07"?

```{r}
summary(vax.36$percent_of_population_fully_vaccinated)
```

> Q18. Using ggplot generate a histogram of this data.

```{r}
ggplot(vax.36) +
  aes(percent_of_population_fully_vaccinated) +
  geom_histogram(bins = 20) +
  xlim(0,1) +
  geom_vline(xintercept = 0.7215, col = "red", linetype = 5)
```

> Q19. Is the 92109 and 92040 ZIP code areas above or below the average value you calculated for all these above?

```{r}
vax %>% filter(as_of_date == "2023-03-07") %>%  
  filter(zip_code_tabulation_area== "92040") %>%
  select(percent_of_population_fully_vaccinated)
```

```{r}
filter(vax.36, zip_code_tabulation_area %in% c("92109", "92040"))
```

They are both below 0.7215.

> Q20. Finally make a time course plot of vaccination progress for all areas in the full dataset with a age5_plus_population \> 36144.

```{r}
last <- vax[vax$age5_plus_population > 36144,]
head(last)
```

```{r}
vax.36.all <- filter(vax, age5_plus_population > 36144)


ggplot(vax.36.all) +
  aes(as_of_date,
      percent_of_population_fully_vaccinated, 
      group=zip_code_tabulation_area) +
  geom_line(alpha=0.2, color="blue") +
  ylim(0,1) +
  labs(x="Date", y="Percent of Population Fully Vaccination",
       title="Vaccination Rate across California",
       subtitle="Only areas with a population above 36k are shown") +
  geom_hline(yintercept = 0.7215, linetype=5)
```

> Q21. How do you feel about traveling for Spring Break and meeting for in-person class afterwards?

I feel pretty safe because the vaccination rate is pretty high.
